---
title: "Reflection on Regression and Classification Models with Group Project Data"
subtitle: "Analytics Objective: Predict Retail Sales Revenue"
author: "Luis Berumen"
date: today
format:
  html:
    toc: true
    toc-depth: 4
    toc-expand: 1
    toc-location: right-body
    toc-title: "Contents"
    number-sections: true
    code-fold: show
    code-overflow: wrap
    code-tools: true
    theme: cosmo
    highlight-style: github
    df-print: kable
    embed-resources: true
    smooth-scroll: true
execute:
  warning: false
  message: false
  freeze: false
editor: visual
---

```{r}
#| label: global-setup
#| include: false

library(tidyverse)
library(janitor)
library(lubridate)
library(tidymodels)
library(glmnet)
library(scales)
library(knitr)
library(patchwork)

tidymodels_prefer()
theme_set(theme_minimal(base_size = 14))
set.seed(2026)

# Locate the retail CSV inside the project folder.
csv_files <- list.files(
  ".",
  pattern = "\\.csv$",
  recursive = TRUE,
  full.names = TRUE
)

data_path <- csv_files[grepl("retail|sales", csv_files, ignore.case = TRUE)][1]
if (is.na(data_path)) data_path <- csv_files[1]
if (is.na(data_path)) stop("No CSV file was found in the project folder.")

retail <- read_csv(data_path, show_col_types = FALSE) |>
  clean_names() |>
  mutate(
    date = as.Date(date),
    month = month(date, label = TRUE, abbr = TRUE),
    year = factor(year(date)),
    season = case_when(
      month(date) %in% c(12, 1, 2) ~ "Winter",
      month(date) %in% c(3, 4, 5)  ~ "Spring",
      month(date) %in% c(6, 7, 8)  ~ "Summer",
      TRUE                         ~ "Fall"
    ),
    season = factor(season, levels = c("Winter", "Spring", "Summer", "Fall")),
    holiday = factor(
      if_else(as.logical(holiday_effect), "Holiday", "Non-holiday")
    ),
    day = factor(
      day_of_the_week,
      levels = c(
        "Monday", "Tuesday", "Wednesday", "Thursday",
        "Friday", "Saturday", "Sunday"
      )
    ),
    discount_group = case_when(
      discount_percentage == 0  ~ "No discount",
      discount_percentage <= 10 ~ "Low (1–10%)",
      TRUE                      ~ "High (11–20%)"
    ),
    discount_group = factor(
      discount_group,
      levels = c("No discount", "Low (1–10%)", "High (11–20%)")
    ),
    product_category = factor(product_category)
  ) |>
  select(
    sales_revenue_usd,
    discount_percentage,
    units_sold,
    marketing_spend_usd,
    product_category,
    day,
    holiday,
    month,
    season,
    year
  )

reg_metrics <- metric_set(rmse, rsq, mae)
```

# Introduction

## Prompt: State your Analytics Objective for the group project. If it did not involve machine learning, revise it so that it is appropriately stated for a machine-learning task.

### Response

My revised analytics objective is:

> **Predict sales revenue from discount percentage and other relevant retail characteristics, compare multiple regression models, and identify the features that contribute most to prediction.**

The original project objective focused on examining the relationship between discounts and revenue. For this assignment, I revised it into a prediction-focused objective because machine learning requires a clearly defined target that the model will predict.

::: {.callout-note title="Customer-oriented business purpose"}
The intended audience is a retail or ecommerce manager who needs an evidence-based way to estimate revenue under different promotional conditions. A useful model can support promotion planning, revenue forecasting, and decisions about when discounts should be paired with marketing support.
:::

This report follows the regression workflow demonstrated in Module 4: data splitting, recipe creation, model specification, fitting, evaluation, cross-validation, hyperparameter tuning, and feature-importance analysis.[^module]

[^module]: The structure is adapted to the Group 1 retail sales dataset rather than the customer-lifetime-value example used in the course codebook.

# Machine-Learning Task and Variable Roles

## Prompt: State whether the objective uses supervised or unsupervised learning.

### Response

This is a **supervised-learning** task because the dataset contains a known outcome—`sales_revenue_usd`—and the models learn from labeled historical observations to predict that outcome for unseen observations.

Unsupervised methods such as clustering would instead search for hidden group structure without a labeled outcome. That is not the primary goal of this analytics objective.

## Prompt: What variables are included, and what roles do they have?

### Response

| Role | Variable | Reason for inclusion |
|---|---|---|
| Outcome | `sales_revenue_usd` | Continuous dollar amount to predict |
| Primary feature | `discount_percentage` | Central promotional variable in RO3 |
| Feature | `units_sold` | Captures transaction volume |
| Feature | `marketing_spend_usd` | Represents promotional support |
| Feature | `product_category` | Allows revenue patterns to vary by product type |
| Feature | `day` | Controls for day-of-week purchasing patterns |
| Feature | `holiday` | Controls for holiday demand |
| Feature | `month`, `season`, `year` | Controls for time and seasonal patterns |

`store_id` and `product_id` are excluded because they function primarily as identifiers. `store_location` is also excluded from the main model because it contains many levels and could create a large, sparse collection of dummy variables that may distract from the promotion-focused objective.

::: {.callout-warning title="Prediction timing and possible leakage"}
`units_sold` is strongly connected to revenue and may only be known after a transaction occurs. Therefore, the main model predicts transaction-level revenue with all available transaction information. A sensitivity model without `units_sold` is also included to show how performance changes when the model is used earlier in the planning process.
:::

## Prompt: Name the broader category of machine-learning methods.

### Response

The broader method is **regression** because the outcome variable is continuous. Classification methods would be appropriate only if the target were categorical, such as high versus low revenue.

# Data Preparation

## Prompt: Set up the QMD file with appropriate libraries and data.

### Response

The global setup chunk loads the tidyverse, tidymodels, glmnet, scales, knitr, and patchwork packages. The retail CSV is located automatically, imported, cleaned with `clean_names()`, and transformed using the same date, holiday, season, and discount logic used in the earlier Group 1 proposal.

## Data overview

```{r}
#| label: tbl-data-overview
#| tbl-cap: "Analysis dataset overview"

retail |>
  summarise(
    observations = n(),
    outcome_missing = sum(is.na(sales_revenue_usd)),
    average_revenue = mean(sales_revenue_usd, na.rm = TRUE),
    median_revenue = median(sales_revenue_usd, na.rm = TRUE),
    average_discount = mean(discount_percentage, na.rm = TRUE),
    average_units = mean(units_sold, na.rm = TRUE),
    average_marketing_spend = mean(marketing_spend_usd, na.rm = TRUE)
  ) |>
  mutate(
    average_revenue = dollar(average_revenue),
    median_revenue = dollar(median_revenue),
    average_discount = percent(average_discount / 100, accuracy = 0.1),
    average_units = round(average_units, 2),
    average_marketing_spend = dollar(average_marketing_spend)
  ) |>
  kable()
```

Table @tbl-data-overview confirms the sample size, outcome completeness, and central values used in the modeling workflow.

## Exploratory relationship

```{r}
#| label: fig-discount-revenue
#| fig-cap: "Discount percentage and sales revenue"
#| fig-height: 5

retail |>
  ggplot(aes(discount_percentage, sales_revenue_usd)) +
  geom_jitter(alpha = 0.10, width = 0.25, color = "#23351F") +
  geom_smooth(method = "lm", se = FALSE, color = "#D9A441", linewidth = 1) +
  scale_x_continuous(labels = label_percent(scale = 1)) +
  scale_y_continuous(labels = label_dollar(scale_cut = cut_short_scale())) +
  labs(
    x = "Discount percentage",
    y = "Sales revenue (USD)"
  )
```

Figure @fig-discount-revenue provides an initial descriptive view. The predictive models below test whether discount percentage adds useful predictive information after accounting for transaction volume, marketing spend, product category, and timing variables.

# Train-Test Split and Preprocessing

## Prompt: Go through the machine-learning process, including data splitting and recipe creation.

### Response

I reserve 20% of the observations as a final test set. The test set is not used for model selection or tuning. Ten-fold cross-validation is performed only within the training set.

```{r}
#| label: data-split

set.seed(2026)
retail_split <- initial_split(retail, prop = 0.80, strata = sales_revenue_usd)
retail_train <- training(retail_split)
retail_test  <- testing(retail_split)

split_summary <- tibble(
  sample = c("Training", "Test"),
  observations = c(nrow(retail_train), nrow(retail_test)),
  mean_revenue = c(
    mean(retail_train$sales_revenue_usd),
    mean(retail_test$sales_revenue_usd)
  )
) |>
  mutate(mean_revenue = dollar(mean_revenue))

kable(split_summary)
```

```{r}
#| label: shared-recipe

retail_rec <- recipe(sales_revenue_usd ~ ., data = retail_train) |>
  step_impute_median(all_numeric_predictors()) |>
  step_impute_mode(all_nominal_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_normalize(all_numeric_predictors()) |>
  step_zv(all_predictors())

retail_rec
```

The same shared recipe is used for OLS, Ridge, Lasso, and Elastic Net so that differences in performance reflect the models rather than inconsistent preprocessing.

```{r}
#| label: cv-folds

set.seed(2026)
retail_folds <- vfold_cv(retail_train, v = 10, strata = sales_revenue_usd)
```

# Initial Method: Elastic Net

## Prompt: First, pick a machine-learning method and explain why the method would be the best.

### Response

I selected **Elastic Net regression** as the initial candidate because this retail dataset includes predictors that may be correlated. For example, units sold, marketing spend, discount percentage, product category, and season may overlap in the information they provide about revenue. Elastic Net combines Ridge shrinkage with Lasso variable selection, making it a reasonable starting method when some predictors may be redundant but still contain useful signal.

The Elastic Net objective combines the residual sum of squares with both L1 and L2 penalties:

$$
\text{Loss} = RSS + \lambda\left[\frac{1-\alpha}{2}\sum_j\beta_j^2 + \alpha\sum_j|\beta_j|\right]
$$

Here, $\lambda$ controls the overall penalty and $\alpha$ controls the mixture of Ridge and Lasso behavior.

## Model specification, fitting, and evaluation

```{r}
#| label: initial-enet

initial_enet_spec <- linear_reg(penalty = 0.1, mixture = 0.5) |>
  set_engine("glmnet")

initial_enet_wf <- workflow() |>
  add_recipe(retail_rec) |>
  add_model(initial_enet_spec)

initial_enet_fit <- fit(initial_enet_wf, data = retail_train)
initial_enet_preds <- augment(initial_enet_fit, new_data = retail_test)
initial_enet_metrics <- initial_enet_preds |>
  reg_metrics(truth = sales_revenue_usd, estimate = .pred)

initial_enet_metrics |>
  mutate(.estimate = round(.estimate, 3)) |>
  kable()
```

::: {.callout-tip title="How the metrics are interpreted"}
- **RMSE:** typical prediction error in revenue dollars, with larger errors penalized more heavily; lower is better.
- **MAE:** average absolute prediction error in revenue dollars; lower is better.
- **R²:** proportion of outcome variation explained by the model; higher is better.
:::

```{r}
#| label: fig-initial-residuals
#| fig-cap: "Residuals from the initial Elastic Net model"
#| fig-height: 5

initial_enet_preds |>
  mutate(residual = sales_revenue_usd - .pred) |>
  ggplot(aes(.pred, residual)) +
  geom_point(alpha = 0.18, color = "#23351F") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "#D9A441") +
  labs(
    x = "Predicted sales revenue (USD)",
    y = "Residual (actual − predicted)"
  )
```

Figure @fig-initial-residuals is used to check whether residuals appear randomly distributed around zero or whether systematic patterns remain.

# Alternative Regression Methods

## Prompt: What alternative methods can improve the metrics produced above? Try them.

### Response

I compare the initial Elastic Net model with four alternatives:

1. **OLS linear regression** as an interpretable baseline.
2. **Ridge regression** to shrink correlated coefficients while retaining every predictor.
3. **Lasso regression** to shrink coefficients and remove weak predictors.
4. **Polynomial Lasso regression** to capture possible curvature in discount, marketing spend, and units sold while controlling complexity through regularization.

::: {.panel-tabset}

## OLS

```{r}
#| label: ols-model

ols_spec <- linear_reg() |>
  set_engine("lm")

ols_wf <- workflow() |>
  add_recipe(retail_rec) |>
  add_model(ols_spec)

ols_fit <- fit(ols_wf, data = retail_train)
ols_preds <- augment(ols_fit, new_data = retail_test)
ols_metrics <- ols_preds |>
  reg_metrics(truth = sales_revenue_usd, estimate = .pred)

ols_metrics |> kable()
```

## Ridge

```{r}
#| label: ridge-model

ridge_spec <- linear_reg(penalty = 0.1, mixture = 0) |>
  set_engine("glmnet")

ridge_wf <- workflow() |>
  add_recipe(retail_rec) |>
  add_model(ridge_spec)

ridge_fit <- fit(ridge_wf, data = retail_train)
ridge_preds <- augment(ridge_fit, new_data = retail_test)
ridge_metrics <- ridge_preds |>
  reg_metrics(truth = sales_revenue_usd, estimate = .pred)

ridge_metrics |> kable()
```

## Lasso

```{r}
#| label: lasso-model

lasso_spec <- linear_reg(penalty = 0.1, mixture = 1) |>
  set_engine("glmnet")

lasso_wf <- workflow() |>
  add_recipe(retail_rec) |>
  add_model(lasso_spec)

lasso_fit <- fit(lasso_wf, data = retail_train)
lasso_preds <- augment(lasso_fit, new_data = retail_test)
lasso_metrics <- lasso_preds |>
  reg_metrics(truth = sales_revenue_usd, estimate = .pred)

lasso_metrics |> kable()
```

## Polynomial + Lasso

```{r}
#| label: poly-model

poly_rec <- recipe(sales_revenue_usd ~ ., data = retail_train) |>
  step_impute_median(all_numeric_predictors()) |>
  step_impute_mode(all_nominal_predictors()) |>
  step_poly(
    discount_percentage,
    marketing_spend_usd,
    units_sold,
    degree = 2
  ) |>
  step_dummy(all_nominal_predictors()) |>
  step_normalize(all_numeric_predictors()) |>
  step_zv(all_predictors())

poly_wf <- workflow() |>
  add_recipe(poly_rec) |>
  add_model(lasso_spec)

poly_fit <- fit(poly_wf, data = retail_train)
poly_preds <- augment(poly_fit, new_data = retail_test)
poly_metrics <- poly_preds |>
  reg_metrics(truth = sales_revenue_usd, estimate = .pred)

poly_metrics |> kable()
```

:::

# Hyperparameter Tuning

## Prompt: Improve the model by tuning its parameters.

### Response

The manually selected Elastic Net values (`penalty = 0.1`, `mixture = 0.5`) are only starting points. I tune both parameters with a regular grid and select the combination with the lowest cross-validated RMSE.

```{r}
#| label: enet-tuning
#| cache: true

set.seed(2026)

tuned_enet_spec <- linear_reg(
  penalty = tune(),
  mixture = tune()
) |>
  set_engine("glmnet")

tuned_enet_wf <- workflow() |>
  add_recipe(retail_rec) |>
  add_model(tuned_enet_spec)

enet_grid <- grid_regular(
  penalty(range = c(-4, 1)),
  mixture(),
  levels = c(5, 5)
)

enet_tune_results <- tune_grid(
  tuned_enet_wf,
  resamples = retail_folds,
  grid = enet_grid,
  metrics = reg_metrics,
  control = control_grid(save_pred = TRUE)
)

best_enet <- select_best(enet_tune_results, metric = "rmse")
best_enet |> kable()
```

```{r}
#| label: fig-enet-tuning
#| fig-cap: "Elastic Net tuning results across penalty and mixture values"
#| fig-height: 6

autoplot(enet_tune_results) +
  labs(
    title = NULL,
    subtitle = "Ten-fold cross-validation; lower RMSE and MAE are better"
  )
```

```{r}
#| label: final-enet

final_enet_wf <- finalize_workflow(tuned_enet_wf, best_enet)
final_enet_fit <- fit(final_enet_wf, data = retail_train)
final_enet_preds <- augment(final_enet_fit, new_data = retail_test)
final_enet_metrics <- final_enet_preds |>
  reg_metrics(truth = sales_revenue_usd, estimate = .pred)

final_enet_metrics |> kable()
```

# Ten-Fold Cross-Validation Comparison

## Prompt: Compare all methods using 10-fold cross-validation. Interpret the results. Which model is best, and does that make sense?

### Response

Every fixed model is compared using the same 10 folds. The tuned Elastic Net result comes directly from the tuning resamples at its selected parameter combination.

```{r}
#| label: cv-comparison
#| cache: true

cv_metrics <- function(wf, label) {
  fit_resamples(
    wf,
    resamples = retail_folds,
    metrics = reg_metrics,
    control = control_resamples(save_pred = TRUE)
  ) |>
    collect_metrics() |>
    mutate(model = label)
}

ols_cv <- cv_metrics(ols_wf, "OLS Linear")
ridge_cv <- cv_metrics(ridge_wf, "Ridge (fixed)")
lasso_cv <- cv_metrics(lasso_wf, "Lasso (fixed)")
enet_cv <- cv_metrics(initial_enet_wf, "Elastic Net (fixed)")
poly_cv <- cv_metrics(poly_wf, "Polynomial + Lasso")

tuned_enet_cv <- collect_metrics(enet_tune_results) |>
  semi_join(best_enet, by = c("penalty", "mixture")) |>
  mutate(model = "Elastic Net (tuned)")

all_cv <- bind_rows(
  ols_cv,
  ridge_cv,
  lasso_cv,
  enet_cv,
  poly_cv,
  tuned_enet_cv
)
```

```{r}
#| label: tbl-cv-results
#| tbl-cap: "Ten-fold cross-validation performance for all regression models"

all_cv |>
  select(model, .metric, mean, std_err) |>
  mutate(
    mean = round(mean, 4),
    std_err = round(std_err, 4)
  ) |>
  arrange(.metric, desc(.metric == "rsq"), mean) |>
  kable()
```

```{r}
#| label: fig-cv-rmse
#| fig-cap: "Ten-fold cross-validation RMSE comparison"
#| fig-height: 5.5

all_cv |>
  filter(.metric == "rmse") |>
  ggplot(aes(reorder(model, mean), mean)) +
  geom_col(fill = "#23351F", alpha = 0.88) +
  geom_errorbar(
    aes(ymin = mean - std_err, ymax = mean + std_err),
    width = 0.25,
    color = "#D9A441",
    linewidth = 0.9
  ) +
  coord_flip() +
  scale_y_continuous(labels = label_dollar()) +
  labs(
    x = NULL,
    y = "Mean cross-validated RMSE"
  )
```

Figure @fig-cv-rmse identifies the model with the lowest average cross-validated RMSE. Differences should be interpreted together with the standard-error bars. If two models have substantially overlapping error bars, the simpler model may be preferable because its performance is not meaningfully worse.

```{r}
#| label: test-comparison

bind_rows(
  ols_metrics |> mutate(model = "OLS Linear"),
  ridge_metrics |> mutate(model = "Ridge (fixed)"),
  lasso_metrics |> mutate(model = "Lasso (fixed)"),
  initial_enet_metrics |> mutate(model = "Elastic Net (fixed)"),
  poly_metrics |> mutate(model = "Polynomial + Lasso"),
  final_enet_metrics |> mutate(model = "Elastic Net (tuned)")
) |>
  select(model, .metric, .estimate) |>
  pivot_wider(names_from = .metric, values_from = .estimate) |>
  arrange(rmse) |>
  mutate(
    rmse = dollar(rmse),
    mae = dollar(mae),
    rsq = percent(rsq, accuracy = 0.1)
  ) |>
  kable(
    col.names = c("Model", "RMSE", "R²", "MAE")
  )
```

::: {.callout-important title="Interpretation to finalize after rendering"}
After the report renders, identify the lowest-RMSE model in Table @tbl-cv-results and explain whether the result is reasonable. A tuned or regularized model should outperform OLS only when shrinkage, variable selection, or nonlinear terms improve generalization. If the metrics are nearly identical, that would suggest the underlying revenue relationship is already captured well by a simple linear structure.
:::

# Feature Importance

## Prompt: Produce a chart that shows the importance of features. Is the result reasonable? Why or why not?

### Response

For the finalized Elastic Net model, feature importance is represented by the absolute magnitude of standardized coefficients. Because the recipe normalizes numeric predictors, coefficient magnitudes are more comparable than they would be on the original scales.

```{r}
#| label: fig-feature-importance
#| fig-cap: "Feature importance from the tuned Elastic Net model"
#| fig-height: 6

importance_tbl <- final_enet_fit |>
  extract_fit_parsnip() |>
  tidy() |>
  filter(term != "(Intercept)", estimate != 0) |>
  mutate(importance = abs(estimate)) |>
  slice_max(importance, n = 15) |>
  arrange(importance)

importance_tbl |>
  ggplot(aes(importance, reorder(term, importance))) +
  geom_col(fill = "#23351F", alpha = 0.88) +
  labs(
    x = "Absolute standardized coefficient",
    y = NULL
  )
```

The result is reasonable if `units_sold` ranks highly because revenue normally increases with transaction volume. Discount percentage and marketing spend may also matter, although their importance should be interpreted as predictive rather than causal. If `units_sold` dominates the chart, that also reinforces the earlier warning that the strongest model may be most useful for transaction-level estimation rather than advance promotion planning.

# Sensitivity Analysis: Excluding Units Sold

## Prompt: What additional method or analysis can improve understanding of the model?

### Response

I fit a second tuned Elastic Net model without `units_sold`. This answers a more operational question: how well can revenue be predicted before the final number of units sold is known?

```{r}
#| label: no-units-model
#| cache: true

retail_no_units <- retail |> select(-units_sold)

set.seed(2026)
no_units_split <- initial_split(
  retail_no_units,
  prop = 0.80,
  strata = sales_revenue_usd
)
no_units_train <- training(no_units_split)
no_units_test <- testing(no_units_split)

no_units_rec <- recipe(sales_revenue_usd ~ ., data = no_units_train) |>
  step_impute_median(all_numeric_predictors()) |>
  step_impute_mode(all_nominal_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_normalize(all_numeric_predictors()) |>
  step_zv(all_predictors())

set.seed(2026)
no_units_folds <- vfold_cv(no_units_train, v = 10, strata = sales_revenue_usd)

no_units_wf <- workflow() |>
  add_recipe(no_units_rec) |>
  add_model(tuned_enet_spec)

no_units_tune <- tune_grid(
  no_units_wf,
  resamples = no_units_folds,
  grid = enet_grid,
  metrics = reg_metrics
)

no_units_best <- select_best(no_units_tune, metric = "rmse")
no_units_final_wf <- finalize_workflow(no_units_wf, no_units_best)
no_units_fit <- fit(no_units_final_wf, data = no_units_train)
no_units_preds <- augment(no_units_fit, new_data = no_units_test)
no_units_metrics <- no_units_preds |>
  reg_metrics(truth = sales_revenue_usd, estimate = .pred)

bind_rows(
  final_enet_metrics |> mutate(model = "Tuned Elastic Net — with units sold"),
  no_units_metrics |> mutate(model = "Tuned Elastic Net — without units sold")
) |>
  select(model, .metric, .estimate) |>
  pivot_wider(names_from = .metric, values_from = .estimate) |>
  mutate(
    rmse = dollar(rmse),
    mae = dollar(mae),
    rsq = percent(rsq, accuracy = 0.1)
  ) |>
  kable(col.names = c("Model", "RMSE", "R²", "MAE"))
```

A meaningful performance decline without `units_sold` would show that transaction volume carries substantial predictive information. The reduced model may still be more useful for planning because its features are more likely to be available before sales occur.

# Module 4 Questions

## Prompt 1: Ridge vs. Lasso coefficient behavior

### Response

The following models use the same large penalty so their coefficient behavior can be compared directly.

```{r}
#| label: exercise-ridge-lasso

ridge_large_wf <- workflow() |>
  add_recipe(retail_rec) |>
  add_model(
    linear_reg(penalty = 10, mixture = 0) |>
      set_engine("glmnet")
  )

lasso_large_wf <- workflow() |>
  add_recipe(retail_rec) |>
  add_model(
    linear_reg(penalty = 10, mixture = 1) |>
      set_engine("glmnet")
  )

ridge_large_fit <- fit(ridge_large_wf, retail_train)
lasso_large_fit <- fit(lasso_large_wf, retail_train)

coef_comparison <- bind_rows(
  ridge_large_fit |>
    extract_fit_parsnip() |>
    tidy() |>
    mutate(model = "Ridge"),
  lasso_large_fit |>
    extract_fit_parsnip() |>
    tidy() |>
    mutate(model = "Lasso")
) |>
  filter(term != "(Intercept)")

coef_comparison |>
  group_by(model) |>
  summarise(
    zero_coefficients = sum(estimate == 0),
    nonzero_coefficients = sum(estimate != 0),
    .groups = "drop"
  ) |>
  kable()
```

Ridge should retain every predictor while shrinking its coefficient. Lasso can set coefficients exactly to zero, producing a smaller model. For a marketing manager facing 50 or more possible predictors, Lasso may be preferred when the goal is a simpler model that highlights a manageable set of signals.

## Prompt 2: Elastic Net mixing parameter

### Response

```{r}
#| label: exercise-mixtures
#| cache: true

mixture_values <- c(0.25, 0.50, 0.75)

mixture_results <- map_dfr(mixture_values, function(mix) {
  spec <- linear_reg(penalty = 0.1, mixture = mix) |>
    set_engine("glmnet")
  wf <- workflow() |>
    add_recipe(retail_rec) |>
    add_model(spec)
  fit_obj <- fit(wf, retail_train)
  preds <- augment(fit_obj, retail_test)
  nonzero <- fit_obj |>
    extract_fit_parsnip() |>
    tidy() |>
    filter(term != "(Intercept)") |>
    summarise(n = sum(estimate != 0)) |>
    pull(n)
  preds |>
    reg_metrics(truth = sales_revenue_usd, estimate = .pred) |>
    mutate(mixture = mix, nonzero_coefficients = nonzero)
})

mixture_results |>
  filter(.metric == "rmse") |>
  select(mixture, nonzero_coefficients, test_rmse = .estimate) |>
  mutate(test_rmse = dollar(test_rmse)) |>
  kable()
```

As the mixture approaches one, the model behaves more like Lasso and may retain fewer predictors. Elastic Net is preferable to pure Lasso when correlated predictors—such as related seasonal or promotion variables—should be stabilized rather than forcing the model to select only one member of a correlated group.

## Prompt 3: Polynomial degree selection

### Response

```{r}
#| label: exercise-poly-degree
#| cache: true

make_poly_wf <- function(degree) {
  rec <- recipe(sales_revenue_usd ~ ., data = retail_train) |>
    step_impute_median(all_numeric_predictors()) |>
    step_impute_mode(all_nominal_predictors()) |>
    step_poly(
      discount_percentage,
      marketing_spend_usd,
      units_sold,
      degree = degree
    ) |>
    step_dummy(all_nominal_predictors()) |>
    step_normalize(all_numeric_predictors()) |>
    step_zv(all_predictors())

  workflow() |>
    add_recipe(rec) |>
    add_model(lasso_spec)
}

poly_degree_cv <- map_dfr(1:3, function(d) {
  cv_metrics(make_poly_wf(d), paste0("Polynomial degree ", d))
})

poly_degree_cv |>
  filter(.metric == "rmse") |>
  select(model, mean, std_err) |>
  arrange(mean) |>
  kable()
```

If degree three does not improve cross-validated RMSE beyond degree two, the added curvature is not justified. Higher-degree terms increase flexibility and therefore increase overfitting risk unless regularization and cross-validation show that the added complexity improves generalization.

## Prompt 4: Tuning Elastic Net

### Response

The tuned Elastic Net analysis above uses 5 penalty levels and 5 mixture levels. This produces $5 \times 5 = 25$ parameter combinations. With 10 folds, the tuning procedure requires $25 \times 10 = 250$ resampled model fits.

```{r}
#| label: exercise-enet-best

best_enet |>
  mutate(
    penalty = signif(penalty, 4),
    mixture = round(mixture, 2)
  ) |>
  kable()
```

The selected mixture indicates whether the best result is closer to Ridge (`mixture = 0`) or Lasso (`mixture = 1`). The additional complexity is worthwhile only if tuned Elastic Net produces a meaningful RMSE improvement over fixed or tuned Lasso.

## Prompt 5: Business interpretation

### Response

```{r}
#| label: exercise-top-features

importance_tbl |>
  slice_max(importance, n = 10) |>
  select(term, estimate, importance) |>
  arrange(desc(importance)) |>
  mutate(
    direction = if_else(estimate > 0, "Positive", "Negative"),
    estimate = round(estimate, 3),
    importance = round(importance, 3)
  ) |>
  kable(
    col.names = c("Feature", "Coefficient", "Importance", "Direction")
  )
```

The three highest-ranked features should be discussed after rendering the table. For each one, I will explain its direction and whether the pattern makes business sense. An important distinction is that predictive importance does not prove that changing a feature will cause revenue to change.

The most actionable lever should come from a variable the business can influence, such as discount percentage or marketing spend, rather than a variable it merely observes, such as season or realized units sold. Any recommendation must also consider the direction of the coefficient and the possibility that the feature is correlated with other predictors.

# Conclusion

The project uses a complete supervised-regression workflow to predict sales revenue. The analysis begins with an Elastic Net candidate, compares it with OLS, Ridge, Lasso, and polynomial alternatives, tunes the key Elastic Net hyperparameters, and evaluates every model using ten-fold cross-validation and a final held-out test set.

The final recommendation should balance three considerations:

1. Predictive accuracy, particularly RMSE and MAE.
2. Stability across cross-validation folds.
3. Business usefulness and interpretability.

The sensitivity analysis without units sold is especially important because the most accurate model is not necessarily the most useful model for advance promotion planning.

# Appendix

## Reproducibility information

```{r}
#| label: session-information

sessionInfo()
```

## GitHub repository

Replace the placeholder below after pushing the QMD and HTML files:

- [GitHub repository](https://github.com/USERNAME/REPOSITORY)

## Published HTML report

Replace the placeholder below after publishing the report through GitHub Pages:

- [Published GitHub Pages report](https://USERNAME.github.io/REPOSITORY/Luis-Berumen-Regression-Classification-Group-Project.html)

## Course codebook used

- [Module 4 — Regression-Focused Methods](https://jaejungca.github.io/regression-models/m04_regression_models.html)
