---
title: "M06 Reflection Essay: Advanced Data Wrangling"
author: "Luis Berumen"
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 4
    toc-expand: 2
    number-sections: true
    code-link: true
    code-fold: show
    code-tools: true
    code-overflow: wrap
    embed-resources: true
    smooth-scroll: true
output-file: index.html
editor: visual
execute:
  echo: true
  warning: false
  message: false
  freeze: auto
---

```{r}
#| label: setup
#| include: false

required_packages <- c("dplyr", "tibble", "knitr")

missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]

if (length(missing_packages) > 0) {
  stop(
    paste0(
      "Install the following packages before rendering: ",
      paste(missing_packages, collapse = ", "),
      ". Run install.packages(c(",
      paste(sprintf('"%s"', missing_packages), collapse = ", "),
      ")) in the R Console."
    )
  )
}

library(dplyr)
library(tibble)
library(knitr)
```

# Reveal.js Impressions

## Prompt 1

*In Step 1, you learned how to design stunning presentation slides with Quarto. What's your impression of the Revealjs presentation? Describe its capabilities as you learned from the presentations. What are its strengths and weaknesses compared with PPT?*

## Response

My overall impression of Reveal.js is that it is a powerful presentation system for people who work with data, code, and research. It creates slides as web-based HTML files, but the slides still behave like a traditional presentation. A presenter can move through slides with the keyboard, open speaker view, display notes on a second screen, use slide numbers, and organize the deck with a table of contents.

Reveal.js also supports features that make a presentation more engaging. These include incremental bullet points, manual pauses, fragments, fade effects, highlighted text, background colors, transitions, columns, callouts, footnotes, logos, and custom themes. Because the presentation is written in Quarto, R code can be placed directly in the document. The code can read data, calculate statistics, and generate charts or tables while the presentation is rendered.

### Strengths compared with PowerPoint

Reveal.js has several strengths for analytics and research presentations:

1.  **Reproducible analysis.** Code, data, charts, and explanations can be stored in the same file.
2.  **Automatic updating.** When the data or code changes, the presentation can be rendered again.
3.  **Consistency.** YAML settings and CSS can apply a common style across the entire slide deck.
4.  **Version control.** A `.qmd` file is text based, so it works well with Git and GitHub.
5.  **Web publishing.** The presentation can be shared through a link and viewed in a browser.
6.  **Literate programming.** The presentation documents both the analytical process and the findings.

### Weaknesses compared with PowerPoint

Reveal.js also has limitations. It requires knowledge of Markdown, YAML, code chunks, file paths, and sometimes CSS. Package conflicts can stop a presentation from rendering, as I experienced when `haven`, `ggplot2`, and `rlang` were not synchronized. PowerPoint is easier for quickly moving objects, changing layouts visually, and creating slides without code.

PowerPoint is also more familiar to many audiences and organizations. A user can drag images, resize text boxes, and adjust visual elements immediately. In Reveal.js, design changes may require editing code and rendering the file again.

The comparison can be summarized in @tbl-comparison.

```{r}
#| label: tbl-comparison
#| tbl-cap: "Comparison of Reveal.js and PowerPoint"

comparison <- tibble(
  Feature = c(
    "Reproducible code and output",
    "Ease of visual editing",
    "Web publishing",
    "Version control",
    "Learning curve",
    "Data-driven charts and tables"
  ),
  `Reveal.js` = c(
    "Excellent",
    "Moderate",
    "Excellent",
    "Excellent",
    "Higher",
    "Excellent"
  ),
  PowerPoint = c(
    "Limited",
    "Excellent",
    "Moderate",
    "Limited",
    "Lower",
    "Manual or externally linked"
  )
)

kable(comparison)
```

Overall, I see Reveal.js as stronger for analytical, technical, and research presentations, while PowerPoint remains stronger for quick visual editing and general business use. The best choice depends on the audience and the purpose of the presentation.

# My Reveal.js Presentation

## Prompt 2

*In Step 2, you were asked to create a Revealjs presentation. Provide a link to the presentation you published on the web. Add at least one reproducible chart or table in the presentation. Briefly describe what the presentation is about. What did you learn from the experience of the Revealjs presentation?*

## Published presentation

**Presentation link:** <https://brahhmen55.github.io/Pratice-Presentation/>

## What I learned

The most important lesson was that a Reveal.js presentation is not just a collection of slides. It can function as a reproducible analytical document. The written story, code, and visual results can be connected in one file.

I also learned the importance of project organization. The `.qmd` file, data folder, stylesheet, bibliography, and images must use correct relative paths. Opening the `.Rproj` file first helps RStudio recognize the project folder as the working directory.

Another lesson was that package management matters. My first render failed because an older version of `rlang` was loaded while `haven` required a newer version. Updating the package and restarting R solved the issue. This experience showed me that reproducibility includes not only code and data, but also the software environment.

Finally, I learned that clear storytelling is more important than adding every possible effect. Transitions and fragments can support the presentation, but they should not distract from the business problem, analytics objective, methods, findings, and implications.

## Main takeaway

Reveal.js is most effective when technical features support a clear analytical story rather than becoming the main focus of the presentation.

# Advanced Data Wrangling

## Prompt 3

*My advanced wrangling lecture in Step 3 introduced new concepts, including how to join relational databases. Pick three different functions, each from a different category. Describe when to use them and give one coding example along with an output for each function. To clarify, you are expected to do literate programming to weave your narratives with the code.*

The three functions selected for this reflection are:

- `left_join()` from relational data and mutating joins
- `across()` from multi-column operations
- `case_when()` from conditional recoding

# `left_join()`

## Category and purpose

`left_join()` belongs to the category of **relational data and mutating joins**. It is used when information is stored in two tables and the analyst wants to add matching columns from the second table to the first table.

A left join preserves every row from the table on the left. When no matching key exists in the right table, the new fields contain `NA`. This behavior is useful when the first table represents the main population that should not lose any records.

For example, one table may contain credit-union members and another may contain current loans. The shared key is `member_id`.

```{r}
#| label: tbl-left-join-input
#| tbl-cap: "Member and loan tables before the join"

members <- tibble(
  member_id = c(101, 102, 103, 104),
  member_name = c("Ana", "Ben", "Carlos", "Diana")
)

loans <- tibble(
  member_id = c(101, 103, 104),
  loan_balance = c(5000, 8200, 3000)
)

members
loans
```

The following code joins the loan information to the member table.

```{r}
#| label: tbl-left-join-output
#| tbl-cap: "Output produced by left_join()"

member_loans <- members |>
  left_join(loans, by = "member_id")

kable(member_loans)
```

The output retains all four members. Ben remains in the data even though no matching loan record exists, so his `loan_balance` is displayed as `NA`. This is preferable to an inner join when the goal is to preserve all members.

::: callout-important
## Key consideration

Before joining, the analyst should verify that the key has the correct type in both tables and determine whether duplicate keys are expected. Duplicate keys can unexpectedly increase the number of rows.
:::

# `across()`

## Category and purpose

`across()` belongs to the category of **applying an operation to multiple columns**. It is commonly used inside `mutate()` or `summarise()`.

Without `across()`, an analyst might repeat the same function separately for many survey variables. With `across()`, the function is written once and applied to a selected group of columns.

The following example contains three employee-rating items measured on a five-point scale.

```{r}
#| label: tbl-across-input
#| tbl-cap: "Employee-rating items"

ratings <- tibble(
  respondent = 1:5,
  q3 = c(4, 5, 3, 4, 5),
  q4 = c(5, 4, 4, 4, 5),
  q5 = c(4, 5, 3, 5, 4)
)

kable(ratings)
```

The next code calculates both the mean and standard deviation for all three rating columns.

```{r}
#| label: tbl-across-output
#| tbl-cap: "Summary statistics generated with across()"

rating_summary <- ratings |>
  summarise(
    across(
      q3:q5,
      list(
        mean = ~ mean(.x, na.rm = TRUE),
        sd = ~ sd(.x, na.rm = TRUE)
      ),
      .names = "{.col}_{.fn}"
    )
  )

kable(rating_summary, digits = 2)
```

This produces six summary columns without repeating the calculation for each variable. The `.names` argument creates clear names such as `q3_mean` and `q3_sd`.

`across()` is especially useful with large survey datasets because many items often require the same recoding, cleaning, or summary operation.

# `case_when()`

## Category and purpose

`case_when()` belongs to the category of **conditional transformation and recoding**. It is useful when a new variable depends on several logical conditions.

Unlike a simple two-category `if_else()`, `case_when()` can evaluate multiple conditions in order. The final `TRUE` condition acts as a default category.

The following example classifies credit-union members by loan balance.

```{r}
#| label: tbl-case-when-input
#| tbl-cap: "Member loan balances before recoding"

balances <- tibble(
  member = c("A", "B", "C", "D", "E"),
  loan_balance = c(200, 1500, 4800, 6000, 12000)
)

kable(balances)
```

The next code creates a categorical variable called `balance_group`.

```{r}
#| label: tbl-case-when-output
#| tbl-cap: "Loan-balance categories produced with case_when()"

balance_groups <- balances |>
  mutate(
    balance_group = case_when(
      loan_balance < 1000 ~ "Low",
      loan_balance < 5000 ~ "Moderate",
      TRUE ~ "High"
    )
  )

kable(balance_groups)
```

The conditions are evaluated from top to bottom. A balance below 1,000 is classified as low. A remaining balance below 5,000 is moderate. All remaining values are classified as high.

::: callout-caution
## Order matters

The conditions in `case_when()` must be arranged carefully. If a broad condition is listed before a narrower condition, the broad condition may classify a value before the narrower rule is reached.
:::

# Function Comparison

The three functions solve different data-wrangling problems, as shown in @tbl-functions.

```{r}
#| label: tbl-functions
#| tbl-cap: "Summary of the three advanced wrangling functions"

function_summary <- tibble(
  Function = c("left_join()", "across()", "case_when()"),
  Category = c(
    "Relational data and joins",
    "Multi-column operations",
    "Conditional recoding"
  ),
  Best_use = c(
    "Adding matching information from another table while preserving all rows in the first table",
    "Applying the same calculation or transformation to several columns",
    "Creating a variable from multiple logical conditions"
  ),
  Main_risk = c(
    "Incorrect or duplicate keys",
    "Selecting unintended columns",
    "Placing conditions in the wrong order"
  )
)

kable(function_summary)
```
